---
title: "Microbial Community Profiling with Differential Analysis"
author: "Bonface Gichuki"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

## Overview

This practical introduces students to taxonomic microbiome analysis using sequencing-derived abundance tables (for example from **Sylph**), focusing on:
  - Community structure
  - Diversity metrics
  - Differential abundance and prevalence

The workflow uses:
- **phyloseq** for data handling and diversity analysis
- **microViz** for transformation and ordination
- **MaAsLin3** for statistical modelling

## Questions addressed in this session

1. How does alpha diversity differ between children with severe malaria (SM) and Community Controls (CC)?
2. Do samples cluster by clinical group in beta diversity (ordination space)?
3. Which taxa dominate the microbial community at different taxonomic levels?
4. Are there community-wide shifts associated with disease status?
5. Which bacterial taxa are differentially abundant between SM and CC?
6. Do results highlight biologically relevant taxa (e.g. Enterobacteriaceae species)?

## Expected learning points
- Taxonomic profiles describe who is present, not just function
- Alpha diversity reflects within-sample diversity
- Beta diversity reflects between-sample differences
- Ordination methods (MDS (Multidimensional Scaling) ) reveal clustering patterns
- Differential prevalence and abundance requires careful statistical modelling
- Compositionality must be considered (e.g. Centered Log-Ratio (CLR) transformation)

## Prerequisites

This module assumes:

- Basic familiarity with R / RStudio
- A Sylph-derived abundance table
- Sample metadata
- Required packages installed

If needed, install missing packages first:
- install.packages("BiocManager")
- BiocManager::install("phyloseq")
- .libPaths("~/R/library")

```{r setup, message=FALSE, warning=FALSE}
# ============================================
# LIBRARIES
# ============================================
library(readr)
library(tidyverse)
library(stringr)
library(tibble)
library(data.table)
library(phyloseq)
library(microbiome)
library(ggpubr)
library(microViz)
library(corncob)
library(ggthemes)
library(gghalves)
library(microbiomeutilities)
library(maaslin3)
```


```{r read_data}
# ============================================
# READ DATA
# ============================================
abundance_data <- read.csv("/Users/LuicerAnneIngasia/Desktop/wcs/sylph_read_based.csv", check.names = FALSE)

metadata <- read.csv("/Users/LuicerAnneIngasia/Desktop/wcs/Bednarski_2025_cleaned_metadata_final.csv", check.names = FALSE) %>%
  column_to_rownames("metagm_id")

# Inspect data
head(abundance_data[, 1:min(6, ncol(abundance_data))])
head(metadata)

# Check overlap between abundance table samples and metadata row names
colnames(metadata)
rownames(metadata)[1:10]
colnames(abundance_data)[1:10]

common_samples_check <- intersect(colnames(abundance_data)[-1], rownames(metadata))
length(common_samples_check)
head(common_samples_check)
```

```{r make-phyloseq}
# ============================================
# CREATE A PHYLOSEQ OBJECT
# ============================================

# First column contains taxonomic labels
otu_tab <- abundance_data %>%
  column_to_rownames("clade_name")

# Convert to matrix
otu_mat <- as.matrix(otu_tab)

# Keep only species-level taxa
otu_mat_species <- otu_mat[grepl("\\|s__[^|]+$", rownames(otu_mat)), , drop = FALSE]

# Align samples between abundance matrix and metadata
common_samples <- intersect(colnames(otu_mat_species), rownames(metadata))
length(common_samples)
head(common_samples)

otu_mat_species <- otu_mat_species[, common_samples, drop = FALSE]
metadata_sub <- metadata[common_samples, , drop = FALSE]

# Sanity check
stopifnot(all(colnames(otu_mat_species) == rownames(metadata_sub)))

# Build taxonomy table from clade strings
tax_split <- strsplit(rownames(otu_mat_species), "\\|")
max_len <- max(lengths(tax_split))

tax_mat <- do.call(rbind, lapply(tax_split, function(x) {
  c(x, rep(NA, max_len - length(x)))
}))

colnames(tax_mat) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "Strain")[1:max_len]
rownames(tax_mat) <- rownames(otu_mat_species)

# Create phyloseq object
ps <- phyloseq(
  otu_table(otu_mat_species, taxa_are_rows = TRUE),
  sample_data(metadata_sub),
  tax_table(as.matrix(tax_mat))
)

# Remove samples with missing disease label
ps <- subset_samples(ps, !is.na(sick))

# Ensure disease group is a factor
sample_data(ps)$sick <- factor(sample_data(ps)$sick, levels = c("CC", "SM"))

# Inspect object
ps
nsamples(ps)
ntaxa(ps)
sample_variables(ps)
sample_sums(ps)[1:min(10, nsamples(ps))]
```

## Community composition: sample-level summary

```{r sample-level-composition, fig.width=10, fig.height=6}
# ============================================
# COMMUNITY COMPOSITION (SAMPLE-LEVEL SUMMARY)
# ============================================

# Clean Family names
tax_table(ps)[, "Family"] <- sub("^f__", "", tax_table(ps)[, "Family"])

# Melt phyloseq object
df <- psmelt(ps)
df$Family <- sub("^f__", "", df$Family)

# Identify top 8 most abundant families
top_families <- df %>%
  filter(!is.na(Family), Family != "") %>%
  group_by(Family) %>%
  summarise(total_abundance = sum(Abundance), .groups = "drop") %>%
  arrange(desc(total_abundance)) %>%
  slice_head(n = 8) %>%
  pull(Family)

top_families

# Collapse everything else into "Other"
df_plot <- df %>%
  mutate(Family_plot = ifelse(Family %in% top_families, Family, "Other"))

# Order samples by sick group
sample_order <- sample_data(ps) %>%
  data.frame() %>%
  rownames_to_column("Sample") %>%
  arrange(sick) %>%
  pull(Sample)

df_plot$Sample <- factor(df_plot$Sample, levels = sample_order)

# Define colours
families_colors <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c",
                     "#fb9a99","#e31a1c","#fdbf6f","#ff7f00","grey")
families_colors <- setNames(families_colors, c(top_families, "Other"))

# Plot
rel_ab_plot <- ggplot(df_plot, aes(x = Sample, y = Abundance, fill = Family_plot)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(~ sick, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = families_colors, name = "Family") +
  labs(x = "Participant", y = "Relative abundance (%)") +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 6, angle = 70, hjust = 1, colour = "black"),
    axis.text.y = element_text(size = 9, colour = "black"),
    axis.title = element_text(size = 9, colour = "black"),
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 7, face = "bold"),
    strip.background = element_blank(),
    strip.text = element_text(size = 10, face = "bold")
  )

rel_ab_plot
```



```{r save-sample-composition, eval=FALSE}
ggsave("sample_rel_ab_plot.png", rel_ab_plot, width = 20, height = 13, units = "cm", dpi = 300)
```

**Interpretation**
The gut microbiome varies substantially between children, but a consistent set of dominant families can be seen across samples. Families such as Bacteroidaceae, Lachnospiraceae, Ruminococcaceae, and Enterobacteriaceae contribute substantially to the community structure.

## Community composition: group-level summary

```{r group-level-composition, fig.width=6, fig.height=5}
# ============================================
# COMMUNITY COMPOSITION (GROUP-LEVEL SUMMARY)
# ============================================

ps_family <- tax_glom(ps, taxrank = "Family")
df_family <- psmelt(ps_family)

df_family$Family <- sub("^f__", "", df_family$Family)
df_family$Family <- ifelse(df_family$Family %in% top_families, df_family$Family, "Other")

df_summary <- df_family %>%
  group_by(sick, Family) %>%
  summarise(mean_abundance = mean(Abundance), .groups = "drop")

rel_ab_group_plot <- ggplot(df_summary, aes(x = sick, y = mean_abundance, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = families_colors) +
  labs(x = "Group", y = "Relative abundance (%)") +
  theme_classic() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9, face = "bold")
  )

rel_ab_group_plot
```


```{r save-group-level-composition, eval=FALSE}
ggsave("group_rel_ab_plot.png", rel_ab_group_plot, width = 8, height = 9, units = "cm", dpi = 300)
```
**Interpretation**
When samples are averaged by clinical group, the relative contribution of major families differs between SM and CC, suggesting broad community restructuring associated with disease status.

## Alpha diversity
```{r alpha-diversity, fig.width=10, fig.height=5}
# ============================================
# ALPHA DIVERSITY
# ============================================

ps_fixed <- ps %>% tax_fix()

alpha_stat_met <- ps_fixed %>%
  ps_calc_diversity(rank = "Species", index = "shannon") %>%
  ps_calc_richness(rank = "Species", index = "observed") %>%
  samdat_tbl()

alpha_stat_met$sick <- factor(alpha_stat_met$sick, levels = c("CC", "SM"))
my_comparisons <- list(c("CC", "SM"))

# Shannon diversity plot
alpha_div_plot_shannon <- ggplot(alpha_stat_met, aes(x = sick, y = shannon_Species, fill = sick)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.8) +
  theme_classic(base_size = 12) +
  labs(x = "Group", y = "Shannon index") +
  scale_fill_manual(values = c("CC" = "#F8766D", "SM" = "#00BFC4")) +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "wilcox.test",
    label = "p.format",
    paired = FALSE
  )

# Observed richness plot
alpha_div_plot_richness <- ggplot(alpha_stat_met, aes(x = sick, y = observed_Species, fill = sick)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.8) +
  theme_classic(base_size = 12) +
  labs(x = "Group", y = "Observed richness") +
  scale_fill_manual(values = c("CC" = "#F8766D", "SM" = "#00BFC4")) +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "wilcox.test",
    label = "p.format",
    paired = FALSE
  )
# Combined figure only
ggpubr::ggarrange(alpha_div_plot_shannon, alpha_div_plot_richness, ncol = 2, nrow = 1)
```


```{r save-alpha-diversity, eval=FALSE}
ggsave("alpha_diversity_shannon_plot.png", alpha_div_plot_shannon, width = 5, height = 5, dpi = 600)
ggsave("alpha_diversity_richness_plot.png", alpha_div_plot_richness, width = 5, height = 5, dpi = 600)
```

**Interpretation**
Both Shannon diversity and observed richness are lower in SM than in CC, indicating that severe malaria is associated with reduced within-sample diversity and fewer detected taxa.

## Beta diversity
```{r beta-diversity, fig.width=5, fig.height=5}
# ============================================
# BETA DIVERSITY + PERMANOVA
# ============================================

ps_species <- ps %>%
  tax_fix() %>%
  tax_filter(min_prevalence = 1/100, use_counts = FALSE, verbose = FALSE) %>%
  tax_agg(rank = "Species")

ps_clr <- ps_species %>%
  tax_transform("clr", zero_replace = "halfmin")

# Ordination
beta_ordination <- ps_clr %>%
  dist_calc("euclidean") %>%
  ord_calc(method = "PCoA") %>%
  ord_plot(color = "sick", alpha = 0.6, size = 2) +
  theme_bw() +
  coord_fixed(0.7) +
  scale_color_manual(values = c("CC" = "#F8766D", "SM" = "#00BFC4")) +
  labs(color = "Group")

# PERMANOVA
perm_vars <- c("sick", "site", "age", "stool_day")
perm_vars <- perm_vars[perm_vars %in% sample_variables(ps_clr)]

perm_res <- ps_clr %>%
  dist_calc("euclidean") %>%
  dist_permanova(
    variables = perm_vars,
    n_perms = 999,
    seed = 123
  ) %>%
  perm_get()

perm_res

# Annotate plot with PERMANOVA result for sick
if ("sick" %in% rownames(perm_res)) {
  pval <- perm_res["sick", "Pr(>F)"]
  r2   <- perm_res["sick", "R2"]

  label_text <- paste0("PERMANOVA\nR² = ", round(r2, 3), "\np = ", signif(pval, 3))

  beta_div_plot <- beta_ordination +
    annotate(
      "text",
      x = Inf, y = -Inf,
      label = label_text,
      hjust = 1.1, vjust = -0.5,
      size = 3
    ) +
    labs(caption = NULL)
} else {
  beta_div_plot <- beta_ordination
}

beta_div_plot
```

```{r save-beta-diversity, eval=FALSE}
ggsave("beta_diversity_plot.png", beta_div_plot, width = 5, height = 5, dpi = 600)
```
**Interpretation**
The PCoA suggests separation between SM and CC in ordination space. PERMANOVA confirms that disease status is significantly associated with overall microbiome composition, although the effect size is modest.

## Differential abundance analysis with MaAsLin3
```{r maaslin3-analysis}
# ============================================
# DIFFERENTIAL ABUNDANCE
# ============================================

# Reuse already loaded objects
sylph_data <- abundance_data
study_metadata <- metadata

# Extract taxa and abundance matrix
taxa <- sylph_data[[1]]
abund <- sylph_data[, -1, drop = FALSE]

# Transpose so rows = samples, columns = taxa
abund_t <- as.data.frame(t(abund))
colnames(abund_t) <- taxa

# Check duplicated taxa names
sum(duplicated(taxa))

# Match samples
common_ids <- intersect(rownames(study_metadata), rownames(abund_t))
length(common_ids)
head(common_ids)

metadata_sub <- study_metadata[common_ids, , drop = FALSE]
features <- abund_t[common_ids, , drop = FALSE]

# Check matching order
all(rownames(features) == rownames(metadata_sub))

# Keep only species-level taxa
features_species <- features[, grepl("\\|s__[^|]+$", colnames(features)), drop = FALSE]

# Ensure model variable is a factor
metadata_sub$sick <- factor(metadata_sub$sick, levels = c("CC", "SM"))
```

```{r maaslin3-fit, eval=FALSE}
# Fit MaAsLin3
fit_out <- maaslin3(
  input_data = features_species,
  input_metadata = metadata_sub,
  output = "maaslin3_output",
  fixed_effects = c("sick"),
  normalization = "CLR",
  transform = "NONE",
  min_abundance = 0.001,
  min_prevalence = 0.1,
  warn_prevalence = FALSE,
  cores = 1
)
```


```{r maaslin-results}
# ============================================
# LOAD AND FILTER RESULTS
# ============================================

res <- read.delim("maaslin3_output/all_results.tsv")

# Keep only sick comparisons
res_sick <- res %>%
  filter(metadata == "sick")

# Significant results
sig_res <- res_sick %>%
  filter(qval_individual < 0.25) %>%
  arrange(qval_individual)

# Clean species names
sig_res$species <- sub(".*\\|s__", "", sig_res$feature)

# Inspect
head(sig_res)
nrow(sig_res)
```
Interpretation.
MaAsLin3 identifies individual species whose abundance or prevalence is associated with disease status after filtering and compositional normalization. Positive coefficients indicate enrichment in SM, while negative coefficients indicate relative depletion in SM.


## Visualise effect sizes (this is what you’re missing)
```{r maaslin-plot, fig.width=6, fig.height=5}
### Option A: simple effect plot
ggplot(sig_res, aes(x = reorder(species, coef), y = coef, fill = coef > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#00BFC4", "FALSE" = "#F8766D")) +
  labs(
    x = "Species",
    y = "Effect size (SM vs CC)",
    title = "Differentially abundant species"
  ) +
  theme_classic()
```
**Interpretation**
The effect-size plot highlights which significant taxa are enriched or depleted in severe malaria. This helps connect statistical output to biological interpretation.


### Option B: volcano-style plot (better for teaching)
```{r volcano-plot, fig.width=6, fig.height=5}
res_sick$species <- sub(".*\\|s__", "", res_sick$feature)

ggplot(res_sick, aes(x = coef, y = -log10(qval_individual))) +
  geom_point(alpha = 0.6) +
  geom_point(data = sig_res, color = "red") +
  labs(
    x = "Effect size",
    y = "-log10(q-value)",
    title = "MaAsLin3 results"
  ) +
  theme_classic()

```
